Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-21 talk/pepm2011/paper.tex
an attempt to rewrite section 4

\lstset{
  basicstyle=\setstretch{1.1}\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{gray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
}

@@

\begin{figure}
\lstset{
    numbers = right,
    numberstyle = \tiny,
    numbersep = 0pt
}
\begin{lstlisting}[mathescape]
# arguments to the trace: $p_{0}$, $p_{1}$
# inside f: res.add(y)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)

@@

\begin{itemize}
    \item \lstinline{new} creates a new object.
    \item \lstinline{get} reads an attribute of an object.
    \item \lstinline{set} writes to an attribute of an object.
    \item \lstinline{guard_class} precedes an (inlined) method call and is
    followed by the trace of the called method.
    \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
    comparison (``greater than''), respectively.
\end{itemize}

@@

The main insight to improve the code shown in the last section is that objects
in category 1 do not survive very long -- they are used only inside the loop
and there is no other outside reference to them. Therefore the optimizer
identifies objects in category 1 and removes the allocation of these objects,
and all operations manipulating them.

This is a process that is usually called \emph{escape analysis}
\cite{goldberg_higher_1990}. In this paper we will
perform escape analysis by using partial evaluation. The use of partial evaluation is a
bit peculiar in that it receives no static input arguments for the trace,
but it is only used to optimize operations within the trace.

To optimize the trace, it is traversed from beginning to end. Every
operation in the input trace is either removed, or new operations are
produced. Whenever a \lstinline{new} operation is seen, the operation it is
removed optimistically and a \emph{static object}\footnote{Here ``static'' is
meant in the sense of partial evaluation, \ie known at partial evaluation time,
not in the sense of static allocation or static method.} is constructed and
associated with the result variable. The static object describes the shape of
the original object, \ie where the values that would be stored in the fields of
the allocated object come from, as well as the type of the object.

When a \lstinline{set} that writes into a static object is optimized, the
corresponding shape description is updated and the operation is removed. This
means that the operation was done at partial evaluation time. When the
optimizer encounters a \lstinline{get} from a static object, the result is read
from the shape description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be
removed as well, because the shape description stores the type and thus the
outcome of the type check the guard does is statically known. Operations that
have dynamic (\ie non-static) objects as arguments are just left untouched by
the optimizer.

In the example from Section~\ref{sub:example}, the following operations
of Figure~\ref{fig:unopt-trace} (lines 10-17) produce two
static objects, and can be completely removed from the optimized trace:

First, the \lstinline{guard_class} operations can be removed, because the classes of $p_{5}$ and
$p_{6}$ are known to be \lstinline{BoxedInteger}. Second, the \lstinline{get} operations can be removed
and $i_{7}$ and $i_{8}$ are just replaced by $i_{4}$ and -100. The only
remaining operation in the optimized trace would be:

The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized similarly.

So far we have only described what happens when static objects are used in guards and in
operations that read and write fields. When the static
object is used in any other operation, it cannot remain static. For example, when
a static object is stored in a globally accessible place, the object has to
be allocated, as it might live longer than one iteration of the loop and as
arbitrary \lstinline{set} operations could change it due to aliasing. This
means that the static
object needs to be turned into a dynamic one, \ie lifted. This makes it
necessary to put operations into the residual code that allocate the
static object at runtime.

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \lstinline{jump} operation
is optimized. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively). This means that instead of a simple jump,
the following operations are emitted:

Observe how the operations for creating these two instances have been moved to later point in the
trace. 
At first sight, it may look like for these operations we didn't gain much, as
the objects are still allocated in the end. However, our optimizations were still
worthwhile, because some operations that have been performed
on the lifted static objects have been removed (some \lstinline{get} operations
and \lstinline{guard_class} operations).

@@

jump($p_{15}$, $p_{10}$)
\end{lstlisting}
\caption{Resulting Trace After Allocation Removal}
\label{fig:step1}
\end{figure}

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \lstinline{guard_class} operations, from the original seven. The
line numbers are the lines where the operations occurred in the original trace
in Figure~\ref{fig:unopt-trace}.

@@

To optimize the simple traces of the last section, we use online partial
evaluation. The partial evaluator optimizes one operation of a trace at a
time. Every operation in the unoptimized trace is replaced by a list of
operations in the optimized trace. This list is empty if the operation
can be optimized away. The optimization rules can be seen in
Figure~\ref{fig:optimization}.

@@

Note in particular that objects in category 1 (\ie those that do
not escape) are completely removed; moreover, objects in category 2
(\ie escaping) are still partially optimized: all the operations in between the
creation of the object and the point where it escapes that involve the object
are removed.

@@

In addition to the count of operations we also performed time measurements.
All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in
Figure~\ref{fig:times}. In addition to the run times the table also reports the
speedup that PyPy achieves when the optimization is turned on.

>>>>

\lstset{
  basicstyle=\setstretch{1.1}\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{gray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
  numberstyle = \tiny,
  numbersep = 0pt,
}

@@

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right]
# arguments to the trace: $p_{0}$, $p_{1}$
# inside f: res.add(y)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)

@@

\begin{itemize}
    \item \lstinline{new} creates a new object.
    \item \lstinline{get} reads an attribute of an object.
    \item \lstinline{set} writes to an attribute of an object.
    \item \lstinline{guard_class} is a precise type check and precedes an
    (inlined) method call and is followed by the trace of the called method.
    \item \lstinline{int_add} and \lstinline{int_gt} are integer addition and
    comparison (``greater than''), respectively.
\end{itemize}

@@

The main insight to improve the code shown in the last section is that objects
in category 1 do not survive very long -- they are used only inside the loop and
there is no other outside reference to them. The optimizer identifies objects in
category 1 and removes the allocation of these objects, and all operations
manipulating them.

This is a process that is usually called \emph{escape analysis}
\cite{goldberg_higher_1990}. In this paper we will perform escape analysis by
using partial evaluation. The use of partial evaluation is a bit peculiar in
that it receives no static input arguments for the trace, but it is only used to
optimize operations within the trace. This section will give an informal account
of this process by examining the example trace in Figure~\ref{fig:unopt-trace}.
The final trace after optimization can be seen in Figure~\ref{fig:step1} (the
line numbers are the lines of the unoptimized trace where the operation comes
from).

@@

jump($p_{15}$, $p_{10}$)
\end{lstlisting}

\caption{Resulting Trace After Allocation Removal}
\label{fig:step1}
\end{figure}

To optimize the trace, it is traversed from beginning to end and an output trace
is produced at the same time. Every operation in the input trace is either
removed, or put into the output trace. Sometimes new operations need to be
produced as well. The optimizer can only remove operations that manipulate
objects that have been allocated within the trace, all others are copied to the
output trace untouched.

Looking at the example trace of Figure~\ref{fig:unopt-trace}, this is what
happens with the operations in lines 1-9. They are manipulating objects that
existed before the trace because they are passed in as arguments. Therefore the
optimizer just puts them into the output trace.

The following operations (lines 10-17) are more interesting:

When the optimizer encounters a \lstinline{new}, it removes it optimistically,
and assumes that the object is in category 1. When the object escapes later, it
will be allocated at that point. The optimizer needs to keep track
of how the object that the operation creates looks like at various points in
the trace. This is done with the help of a \emph{static object}\footnote{Here
``static'' is meant in the sense of partial evaluation, \ie known at partial
evaluation time, not in the sense of ``static allocation'' or ``static
method''.}. The static object describes the shape of the object that would have
been allocated, \ie the type of the object and where the values that would be
stored in the fields of the allocated object come from.

In the snippet above, the two \lstinline{new} operations are removed and two
static objects are constructed. The \lstinline{set} operations manipulate a
static object, therefore they can be removed as well and their effect is
remembered in the static objects.

The \lstinline{guard_class} operations can be removed, since their argument is a
static object with the matching type \lstinline{BoxedInteger}. The
\lstinline{get} operations can be removed as well, because each of them reads a
field out of a static object. The results of the get operation are replaced with
what the static object stores in these fields: $i_{7}$ and $i_{8}$ are just
replaced by $i_{4}$ and -100. The only operation put into the optimized trace
is the addition:

The rest of the trace from Figure~\ref{fig:unopt-trace} is optimized in a
similar vein. The operations in lines 27-35 produce two more static objects and
are removed. Those in line 36-39 are just put into the output trace because they
manipulate objects that are allocated before the trace. Lines 40-42 are removed
because they operate on a static object. Line 43 is put into the output trace.
Lines 44-46 produce a new static object and are removed, lines 48-51 manipulate
that static object and are removed as well. Lines 52-54 are put into the output
trace.

The last operation (line 55) is an interesting case. It is the \lstinline{jump}
operation that passes control back to the beginning of the trace. The two
arguments to this operation are at this point static objects. However, because
they passed into the next iteration of the loop they live longer than the trace
and therefore cannot remain static. They need to be turned into a dynamic
(runtime) object before the actual \lstinline{jump} operation. This process of
turning a static object into a dynamic one is called \emph{lifting}.

Lifting a static object puts \lstinline{new} and \lstinline{set} operations into
the output trace. Those operations produce an object at runtime that has the
same shape that the static object describes. This process is a bit delicate,
because the static objects could form an arbitrary graph structure. In our
example is is simple, though:

Observe how the operations for creating these two instances have been moved to a
later point in the trace. This is worthwhile even though the objects have to be
allocated in the end because some \lstinline{get} operations and
\lstinline{guard_class} operations on the lifted static objects could be
removed.

A bit more generally, lifting needs to occur if a static object is used in any
operation apart from \lstinline{get}, \lstinline{set}, and \lstinline{guard}.
It also needs to occur if \lstinline{set} is used to store a static object into
a non-static one.

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \lstinline{guard_class} operations, from the original seven.

@@

To optimize the simple traces of the last section, we use online partial
evaluation. The partial evaluator optimizes one operation of a trace at a
time. Every operation in the unoptimized trace is replaced by a list of
operations in the optimized trace. This list is empty if the operation can be
optimized away. The optimization rules can be seen in
Figure~\ref{fig:optimization}. Lists are written using angular brackets $<...>$,
list concatenation is expressed using two colons: $l_1::l_2$.

XXX input/output of optimizer

@@

Note in particular that objects in category 1 (\ie those that do
not escape) are completely removed; moreover, objects in category 2
(\ie escaping) are still partially optimized: all the operations in between the
creation of the object and the point where it escapes that involve the object
are removed. Objects in category 3 and 4 are also partially optimized, their
allocation is delayed till the end of the trace.

@@

In addition to the count of operations we also performed time measurements.
All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in
Figure~\ref{fig:times}. In addition to the run times the table also reports the
speedup that PyPy achieves when the optimization is turned on. XXX sounds sucky
